---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gcvisualyst

<!-- badges: start -->
<!-- badges: end -->

The goal of gcvisualyst is to analyze and visualize the GC content of DNA sequences across sliding windows. This tool provides a simple and effective way to calculate GC content and generate plots that illustrate GC content variations along sequences.

Features:
  
  - *GC Content Calculation*: Computes the GC content of DNA sequences across sliding windows of user-defined size.
  
  - *Data Visualization:* Generates visualizations of GC content for multiple sequences, either combined in a single plot or as separate facets for each sequence.
  
  - *Customizable Plots:* Provides options for combined or facet-wrapped layouts for easy comparison of multiple sequences.
  
  - *Efficient Processing:* Utilizes dplyr, ggplot2, stringr, and purrr for efficient data manipulation and plotting.

## Installation

You can install the development version of gcvisualyst like so:

``` r
# install.packages("devtools")
devtools::install_github("woosflex/gcvisualyst")
```

## Usage

GC Visualyst can be used to generate combined plots in which all sequences are plotted in a single graph by using `combined = TRUE` argument in `gc_visualize()` function:

```{r example}
library(gcvisualyst)
## Demo DNA sequences
sequences_df <- data.frame(
  headers = c(
    "seq1", "seq2"
    ),
  sequences = c(
    "AGCTGCGCGTATCGTACGCGATCGTATCGCGATCGTATCGCG",
    "GGCGCGCTAGCTCGAGTCGCGCGGCTCGATAGCTCGTACGTAG"
  ),
  stringsAsFactors = FALSE
)

## Calculate GC content with a sliding window of 10
gc_content_df <- gc_content(sequences_df, window = 10)

## Visualize GC content
gc_visualize(gc_content_df, combined = TRUE)
```

While the other way to use the the package is to generate different plot for all the sequences by using `combined = FALSE` argument, which is the default parameter of the `combined` argument of `gc_visualize()` function:
```{r}
library(gcvisualyst)
## Demo DNA sequences
sequences_df <- data.frame(
  headers = c(
    "seq1", "seq2"
    ),
  sequences = c(
    "AGCTGCGCGTATCGTACGCGATCGTATCGCGATCGTATCGCG",
    "GGCGCGCTAGCTCGAGTCGCGCGGCTCGATAGCTCGTACGTAG"
  ),
  stringsAsFactors = FALSE
)

## Calculate GC content with a sliding window of 10
gc_content_df <- gc_content(sequences_df, window = 10)

## Visualize GC content
gc_visualize(gc_content_df)
```

